---
title: "p8105_hw3_yj2802"
author: "Yizhen Jia"
date: '`r format(Sys.time(), "%Y-%m-%d")`'
output: github_document

---

```{r setup, include=FALSE}
library(tidyverse)
library(ggridges)
library(patchwork)
library(readr)
library(p8105.datasets)

knitr::opts_chunk$set(
	echo = TRUE,
	warning = FALSE,
	fig.width = 8, 
  fig.height = 6,
  out.width = "90%"
)

theme_set(theme_minimal() + theme(legend.position = "bottom"))

options(
  ggplot2.continuous.colour = "viridis",
  ggplot2.continuous.fill = "viridis"
)

scale_colour_discrete = scale_colour_viridis_d
scale_fill_discrete = scale_fill_viridis_d
```

### Problem 1

#### Read in the data

```{r}
data("ny_noaa")
```

#### Answer questions about the data

This dataset contains `r nrow(ny_noaa)` rows and `r ncol(ny_noaa)` columns. Variables include weather station id, date of observation,  (tenths of mm), snowfall (mm), snow depth (mm), and min and max temperature (tenths of degrees C).

Below we clean the data, creating separate variables for year, month, and day and converting `tmax` and `tmin` to numeric. We find that 0 is the most commonly observed value for snowfall. This is because most days of the year, it does not snow at all in NY. The second most commonly observed value is `NA`, indicating missingness. Other common values are 13, 25, and 51, suggesting that snowfall is originally recorded in fractions of an inch and converted to mm.

```{r}
ny_noaa |>  
  count(snow) |> 
  arrange(desc(n))

ny_noaa = 
  ny_noaa |> 
  separate(date, into = c("year", "month", "day"), convert = TRUE) |> 
  mutate(
    tmax = as.numeric(tmax),
    tmin = as.numeric(tmin))
```

Below is a two-panel plot showing the average max temperature in January and in July in each station across years. As expected, the mean temperature in January is much lower than the mean temperature in July for all stations and across all years. All stations appear to follow similar trends of temperature peaks and valleys within a month across the years, i.e. when one station has a high monthly mean temperature for a given year, most other stations also have a high monthly mean temperature for that year. We do see one uncharacteristically cold station in July of 1987 or 1988, as well as a few other less drastic outliers.

```{r}
ny_noaa |> 
  group_by(id, year, month) |> 
  filter(month %in% c(1, 7)) |> 
  summarize(mean_tmax = mean(tmax, na.rm = TRUE, color = id)) |> 
  ggplot(aes(x = year, y = mean_tmax, group = id)) + geom_point() + geom_path() +
  facet_grid(~month) +
  labs(title = "Mean monthly temperature for each station across years for January and July")
```

Below we show a two-panel plot including (i) a hex plot of `tmax` vs `tmin` for the full dataset; and (ii) a ridge plot showing the distribution of snowfall values (in mm) greater than 0 and less than 100 separately by year. 

From the hex plot we see that while there is some variability, the majority of the data cluster tightly in the center of the distribution. In relatively rare cases, it seems that `tmax` is less than `tmin`, which raises questions about data recording and quality.

From the ridge plot, we see a multimodal density of snowfall within a given year. Most stations see between 0 and  35 mm of snow in a year. Then there is a another group of stations that see about 45 mm of snow, and another group that sees nearly 80 mm. It is likely this multimodality stems from the conversion of measurements in one system (fractions of an inch) to another (using the metric system), which was also noted in the table of common values. 

```{r}
hex = 
  ny_noaa |> 
  ggplot(aes(x = tmin, y = tmax)) + 
  geom_hex()

ridge = 
  ny_noaa |> 
  filter(snow < 100, snow > 0) |>
  ggplot(aes(x = snow, y = as.factor(year))) + 
  geom_density_ridges()

hex + ridge
```

### Problem 2

This problem uses accelerometer data collected on 250 participants in the NHANES study. Two datasets including informaton of the participants’ demographic data their accelerometer data, respectively.

#### First, import the `nhanes_covar.csv` dataset:

The dataset includes SEQN, sex, age, BMI and education status data of 250 participants. Those are missing demographic data and those are less than 21 years old should be excluded; SEQN, sex and education should not be numeric.

```{r clean_covar}
covar = 
  read_csv("C:/Users/22865/OneDrive/Desktop/Data Science 1/HWs/p8105_hw3_yj2802/nhanes_covar.csv", skip = 4,
             na = c("NA"), 
             col_types = cols(SEQN = "c", sex = "c", age = "n", BMI = "n", education = "c")) |>
  janitor::clean_names() |>
  drop_na(seqn, sex, age, bmi, education) |>
  filter(age >= 21)
```

After cleaning the data, the covar dataset now has `r ncol(covar)` variables and `r nrow(covar)` observations.

Now, create a table for the number of men and women in each education category.

```{r education*sex table}
covar = covar |>
  mutate(
    sex = recode(sex, `1` = "Male", `2` = "Female"),
    education = recode(education, 
                       `1` = "Less than high school", 
                       `2` = "High school equivalent", 
                       `3` = "More than high school")
  ) |>
  mutate(
    education = factor(education, 
                       levels = c("Less than high school", 
                                  "High school equivalent", 
                                  "More than high school"),
                       ordered = TRUE)
  )

p2_table = covar |>
  group_by(education, sex) |>
  summarize(count = n(), .groups = 'drop') |>
  pivot_wider(names_from = sex, values_from = count, values_fill = 0) |>
  knitr::kable()

p2_table

```

Then, create a visualization of the age distributions for men and women in each education category

```{r ggplot}
ggplot(covar, aes(x = age, y = education, fill = sex)) +
  geom_density_ridges(alpha = 0.7, scale = 1) +
  facet_wrap(~sex) +
  labs(
    title = "Age Distribution by Education Level for Men and Women",
    x = "Age",
    y = "Education Level"
  ) +
  theme_minimal() +
  scale_fill_manual(values = c("Male" = "#1f77b4", "Female" = "#ff7f0e"))
```

#### Comments on the table and plot:

The table shows that the "More than high school" category has the highest count for both males (56) and females (59), while the "Less than high school" category has nearly equal counts for both genders. The ridge plot reveals that higher education levels correspond to younger age distributions, whereas lower education levels are associated with a broader and older age range.

#### Then, import the `nhanes_accel.csv` dataset and merge 2 datasets:

```{r p2_final}
accel = read_csv("C:/Users/22865/OneDrive/Desktop/Data Science 1/HWs/p8105_hw3_yj2802/nhanes_accel.csv",
             col_types = cols(SEQN = "c")) |>
  janitor::clean_names()

p2_final = covar |> inner_join(accel, by = "seqn")
```

The merged dataset "p2_final" has `r ncol(p2_final)` variables, including demographic and accelerometer data for  `r nrow(p2_final)` observations.

#### Create a total activity variable for each participant:

```{r}
activity_columns = select(p2_final, starts_with("min"))
p2_final = p2_final |>
  mutate(total_activities = rowSums(activity_columns, na.rm = TRUE))
```

#### Plot these total activities against age:
```{r ggplot2}
ggplot(p2_final, aes(x = age, y = total_activities, color = sex)) +
  geom_point(alpha = 0.5) +
  geom_smooth(method = "loess", se = FALSE) +
  facet_wrap(~ education) +
  labs(
    title = "Total Activities vs. Age, Compared by Sex and Education Level",
    x = "Age",
    y = "Total Activities"
  ) +
  theme_minimal() +
  scale_color_manual(values = c("Male" = "#1f77b4", "Female" = "#ff7f0e"))
```

This plot shows the relationship between total daily activities and age, comparing men and women across different education levels. The activity patterns generally decline with age, and men tend to have slightly lower total activities compared to women, particularly in the "More than high school" category.

#### Make a three-panel plot that shows the 24-hour activity time courses for each education level:

```{r three-panel plot}
p2_long = p2_final |>
  pivot_longer(
    cols = starts_with("min"),
    names_to = "minute",
    names_prefix = "min",
    values_to = "activity"
  ) |>
  mutate(minute = as.numeric(minute))
ggplot(p2_long, aes(x = minute, y = activity, color = sex)) +
  geom_line(alpha = 0.7) +
  facet_wrap(~ education) +
  labs(
    title = "24-Hour Activity Time Courses by Education Level",
    x = "Minute of the Day",
    y = "Activity (MIMS)",
    color = "Sex"
  ) +
  theme_minimal() +
  scale_color_manual(values = c("Male" = "#1f77b4", "Female" = "#ff7f0e"))
```

This graph shows a consistent pattern of daily activity peaks for both men and women across all education levels. Men generally exhibit higher activity levels, especially in the "More than high school" category, while women show more sporadic peaks in activity. There is less variation in activity levels at the start and end of the day across all education levels.

### Problem 3

#### Import, clean and combine datasets:
```{r import 4 datasets}
Jan2020 = 
  read_csv("C:/Users/22865/OneDrive/Desktop/Data Science 1/HWs/p8105_hw3_yj2802/citibike/Jan 2020 Citi.csv",
             col_types = cols(duration = "n")) |>
  janitor::clean_names() |>
  mutate(year="2020", month = "Jan")

Jan2024 = 
  read_csv("C:/Users/22865/OneDrive/Desktop/Data Science 1/HWs/p8105_hw3_yj2802/citibike/Jan 2024 Citi.csv",
             col_types = cols(duration = "n")) |>
  janitor::clean_names() |>
  mutate(year="2024", month = "Jan")

July2020 = 
  read_csv("C:/Users/22865/OneDrive/Desktop/Data Science 1/HWs/p8105_hw3_yj2802/citibike/July 2020 Citi.csv",
             col_types = cols(duration = "n")) |>
  janitor::clean_names() |>
  mutate(year="2020", month = "July")

July2024 = 
  read_csv("C:/Users/22865/OneDrive/Desktop/Data Science 1/HWs/p8105_hw3_yj2802/citibike/July 2024 Citi.csv",
             col_types = cols(duration = "n")) |>
  janitor::clean_names() |>
  mutate(year="2024", month = "July")

p3_final = bind_rows(Jan2020, Jan2024, July2020, July2024)
```

The result dataset contains Citi Bike data of 4 months, and has `r ncol(p3_final)` variables and `r nrow(p3_final)` observations. Including virables of `r colnames(p3_final)`

#### Create table 1:

Table 1 shows the total number of rides in each combination of year and month separating casual riders and Citi Bike members:

```{r}
table1 = p3_final |>
  group_by(year, month, member_casual) |>
  summarize(total_rides = n(), .groups = 'drop')
table1
```

#### Create table 2:

Table 2 shows the 5 most popular starting stations for July 2024:

```{r}
table2 = p3_final |>
  mutate(year = as.character(year), month = as.character(month)) |>
  filter(year == "2024", month == "July") |>
  group_by(start_station_name) |>
  summarize(total_rides = n(), .groups = 'drop') |>
  arrange(desc(total_rides)) |>
  head(5)
table2
```

#### Create the plot:

The plot shows the effects of day of the week, month, and year on median ride duration:

```{r p3_plot}
median_duration_data = p3_final |>
  group_by(year, month, weekdays) |>
  summarize(median_duration = median(duration, na.rm = TRUE), .groups = 'drop')

ggplot(median_duration_data, aes(x = weekdays, y = median_duration, color = month, group = interaction(month, year))) +
  geom_line() +
  facet_wrap(~ year) +  # Panel by year
  labs(title = "Relationship Between Median Ride Duration and Day of the Week, Month, and Year",
       x = "Day of the Week", y = "Median Ride Duration (minutes)",
       color = "Month") +
  theme_minimal()
```

#### Create the figure

The p3_figure shows the impact of month, membership status, and bike type on the distribution of ride duration for data in 2024:

```{r}
data_2024 = p3_final |>
  filter(year == "2024")

ggplot(data_2024, aes(x = duration, fill = member_casual)) +
  geom_histogram(binwidth = 5, position = "dodge") +
  facet_grid(month ~ rideable_type) +
  labs(title = "Distribution of Ride Duration by Month, Membership Status, and Bike Type (2024)",
       x = "Ride Duration", 
       y = "Number of Rides",
       fill = "Membership Status") +
  theme_minimal()
```

The plot shows that most Citi Bike rides in both January and July 2024, regardless of bike type or membership status, have relatively short duration, with the majority of rides lasting under 50 minutes. Additionally, members account for a significantly higher number of rides compared to casual riders.
